Comparing and combining network-guided biomarker discovery to discover genetic susceptibility mechanisms to breast cancer on the GENESIS study

### 2021-03-03 manuscript/plos/review_4.tex
minor modifications

We conducted association analyses in the GENESIS dataset (Section~\ref{methods:data}) at both SNP and gene levels (Section~\ref{methods:conventional}). At the SNP level, two genomic regions had a P\=/value lower than the Bonferroni threshold on chromosomes 10 and 16 (\nameref{sfig:snp_gene_manhattan}A). The former overlaps with the gene \emph{FGFR2}, the latter with \emph{CASC16} and the protein-coding gene \emph{TOX3}. Variants in both \emph{FGFR2} and \emph{TOX3} have been repeatedly associated with breast cancer susceptibility in other case-control studies \cite{Michailidou2017}, in studies on \emph{BRCA1} and \emph{BRCA2} carriers \cite{Mulligan2011}, and in hereditary breast and ovarian cancer families negative for mutations in \emph{BRCA1} and \emph{BRCA2} \cite{rinella_genetic_2013}. At the gene level, only \emph{FGFR2} was significantly associated with breast cancer (\nameref{sfig:snp_gene_manhattan}B).

Closer examination revealed two other regions (3p24 and 8q24) having low, albeit not genome-wide significant, P\=/values. Both of them have been associated with breast cancer susceptibility in the past \cite{brisbin_meta-analysis_2011,search_newly_2009}. We applied an L1-penalized logistic regression using all GENESIS genotypes as input and the phenotype (cancer/healthy) as the outcome (Section~\ref{methods:classifier}). The algorithm selected 100 SNPs, both from all regions mentioned above and new ones (\nameref{sfig:snp_gene_manhattan}C). However, it was unclear why those SNPs were selected, as emphasized by the high P\=/value of some of them, which further complicates the biological interpretation. Moreover, and in opposition to what would be expected under the omnigenic model, the genes to which these SNPs map (Section~\ref{methods:snp2gene}) were not interconnected in the protein-protein interaction network (PPIN, Section~\ref{methods:networks}). Moreover, the classification performance of the model was low (sensitivity = 55\%, specificity = 55\%, Section~\ref{methods:benchmark}). Together, these issues motivated exploring network methods, which consider not only statistical association but also the location of each gene in a PPIN to find susceptibility genes.

@@

Topologically, the consensus consisted of a connected component composed of 49 genes and multiple smaller subnetworks (Fig~\ref{fig:consensus}B and C). Among the latter, 19 genes were in subnetworks containing a single gene or two connected nodes. This implied that they did not have a consistently altered neighborhood but were strongly associated themselves and hence picked by at least two methods. The large connected component contained genes that are highly central in the PPIN. This property weakly anticorrelated with the P\=/value of association to the disease (Pearson correlation coefficient~=~-0.26, \nameref{sfig:consensus_stats}). This anticorrelation suggested that these genes were selected because they were on the shortest path between two high scoring genes. Because of this, we hypothesize that highly central genes might contribute to the heritability through alterations of their neighborhood, consistent with the omnigenic model of disease \cite{boyle_expanded_2017}. For instance, the most central node in the consensus solution was \emph{COPS5}, a component of the COP9 signalosome that regulates multiple signaling pathways. \emph{COPS5} is related to multiple hallmarks of cancer and is overexpressed in multiple tumors, including breast and ovarian cancer \cite{liu_jab1_cops5_2018}. Despite its lack of association in GENESIS or in studies conducted by the Breast Cancer Association Consortium (BCAC) \cite{Michailidou2017} (VEGAS2 P\=/value of 0.22 and 0.14, respectively), its neighbors in the consensus solution had consistently low P\=/values (median VEGAS2 P\=/value~=~0.006).

@@

We compared the results obtained with different network methods to the European sample of BCAC, the largest GWAS to date on breast cancer (Section~\ref{methods:bcac}). Although BCAC case-control studies do not necessarily target cases with a familial history of breast cancer like GENESIS does, this comparison is pertinent since we expect a shared genetic architecture at the gene level, at which most network methods operate. Together with BCAC's scale (90 times more samples than GENESIS), this shared genetic architecture provided a reasonable counterfactual of what we would expect if GENESIS had a larger sample size. We computed a gene association score on BCAC (Section~\ref{methods:bcac}). The solutions provided by the different  network methods overlapped significantly with BCAC hits (Fisher's exact test P\=/value \textless~0.019). The gene-based methods achieved comparable precision (2\%-25\%) and recall (1.3-12.1\%) at recovering BCAC-significant genes (\nameref{sfig:additional-benchmarks}A). Interestingly, while SConES GI achieved a similar recall at the SNP-level (8.6\%), it showed a much higher precision (47.3\%).

@@

We compared the six network methods in a 5-fold subsampling setting (Section~\ref{methods:benchmark}). In this comparison we measured properties (Fig~\ref{fig:benchmark} and \nameref{sfig:additional-benchmarks}): the size of the solution; the sensitivity and the specificity of an L1-penalized logistic regression classifier on the selected SNPs; the stability of the methods; and their computational runtime. The solution size varies greatly between the different methods (Fig~\ref{fig:benchmark}A). Heinz produced the smallest solutions, with an average of 182 selected SNPs (Section~\ref{methods:snp2gene}) while the largest ones came from SConES GI (6\,256.6 SNPs) and dmGWAS (4\,255.0 SNPs). LEAN did not produce any solution in any of the subsamples.

To determine whether the selected SNPs could predict cancer susceptibility, we computed the classifiers' performances on \emph{test sets} (\nameref{sfig:additional-benchmarks}B). The different classifiers displayed similarly low sensitivities and specificities, all in the 0.52 -- 0.56 range. Interestingly, the classifier trained on all the SNPs had a similar performance, despite being the only method aiming to minimize prediction error. Of course, although these performances were low, we did not expect to separate cases from controls well using exclusively genetic data \cite{deloscamposComplexTraitPredictionEra2018}.

@@

In the GENESIS dataset, heinz's solution was very conservative, providing a small solution with the lowest median P\=/value (Table~\ref{tab:gene_solutions}). By repeatedly selecting this compact solution, heinz was the most stable method (Fig~\ref{fig:benchmark}B). Its conservativeness stemmed from its preprocessing step, which modeled the gene P\=/values as a mixture model of a beta distribution and a uniform distribution, controlled by an FDR parameter. Due to the limited signal at the gene level in this dataset (\nameref{sfig:snp_gene_manhattan}B), only 36 genes retained a positive score after that transformation. However, this small solution did not provide much insight into the susceptibility mechanisms to cancer. Importantly, it ignored genes that were associated with cancer in this dataset, like \emph{FGFR2}.

@@

We explored methods' parameter space by running them under different combinations of parameters (Section~\ref{methods:parameters}). In agreement with their formulations (Section~\ref{methods:methods}), larger values of specific parameters produced less stringent solutions (\nameref{sfig:params_size}A): for HotNet2 and heinz, this is the threshold above which genes receive a positive score; for dmGWAS, it is the \texttt{d} parameter, which controls how far neighbors could be added; for SigMod, it is \texttt{nmax}, which specifies the maximum size of the solution; and for LEAN, it is the P-value threshold to consider a solution significant. Two parameters had the opposite effect (the larger, the more stringent): SigMod's \texttt{maxjump}, which sets the threshold to consider an increment in $\lambda$ ``large enough''; and SConES' $\eta$, where higher values produce smaller solutions. However, two of the parameters did not have the expected effect: dmGWAS' \texttt{r}, which controls the minimum increment in the score required to add a gene; and SigMod's \texttt{maxjump}, which sets the threshold to consider an increment in $\lambda$ ``large enough''. In both cases, the size of the solution was very similar across the different values. Despite the differences in size, the solutions' size was relatively robust to the choice of parameters (\nameref{sfig:params_size}B).

@@

Most network methods, including the consensus, were highly unstable (Fig~\ref{fig:benchmark}B), raising questions about the results' reliability. We built a new, \emph{stable consensus} solution using the genes selected most often across the 30 solutions obtained by running the 6 methods on 5 different splits of the data (Section~\ref{methods:benchmark}). Such a network should capture the subnetworks more often found altered, and hence should be more resistant to noise. We used only genes selected in at least 7 solutions, which corresponded to 1\% of all genes selected at least once. The resulting stability-based consensus was composed of 68 genes (Fig~\ref{fig:stable-consensus}A). This network shared most of the properties of the consensus: breast cancer susceptibility genes were overrepresented (P\=/value = 3 \texttimes{} 10\textsuperscript{-4}), as well as genes involved in mitochondrial translation and the attenuation phase (adjusted P\=/values 0.001 and 3 \texttimes{} 10\textsuperscript{-5} respectively); the selected genes were more central than average (P\=/value = 1.1 \texttimes{} 10\textsuperscript{-14}); and a considerable number of nodes (19) were isolated (Fig~\ref{fig:stable-consensus}B and C).

@@

\paragraph*{S11 Fig}
\label{sfig:scones_gsm}
\textbf{Overview of the solutions produced by the SConES on the GS and GM networks (Section~\ref{methods:networks}) on the GENESIS dataset.} \textbf{(A)}~Manhattan plots of SNPs (Section~\ref{methods:conventional}); in black, the method’s solution. The red line indicates the Bonferroni threshold (2.54 \texttimes{} 10\textsuperscript{-7}). \textbf{(B)}~Precision and recall of the evaluated methods with respect to Bonferroni-significant SNPs (SConES) or genes (other methods) in BCAC. For reference, we added a gray line with a slope of 1. \textbf{(C)}~Solution networks.
%\includegraphics[width=\linewidth]{./figures/sfigure_11.png}

>>>>

We conducted association analyses in the GENESIS dataset (Section~\ref{methods:data}) at both SNP and gene levels (Section~\ref{methods:conventional}). At the SNP level, two genomic regions had a P\=/value lower than the Bonferroni threshold on chromosomes 10 and 16 (S2A Fig). The former overlaps with the gene \emph{FGFR2}, the latter with \emph{CASC16} and the protein-coding gene \emph{TOX3}. Variants in both \emph{FGFR2} and \emph{TOX3} have been repeatedly associated with breast cancer susceptibility in other case-control studies \cite{Michailidou2017}, in studies on \emph{BRCA1} and \emph{BRCA2} carriers \cite{Mulligan2011}, and in hereditary breast and ovarian cancer families negative for mutations in \emph{BRCA1} and \emph{BRCA2} \cite{rinella_genetic_2013}. At the gene level, only \emph{FGFR2} was significantly associated with breast cancer (S2B Fig).

Closer examination revealed two other regions (3p24 and 8q24) having low, albeit not genome-wide significant, P\=/values. Both of them have been associated with breast cancer susceptibility in the past \cite{brisbin_meta-analysis_2011,search_newly_2009}. We applied an L1-penalized logistic regression using all GENESIS genotypes as input and the phenotype (cancer/healthy) as the outcome (Section~\ref{methods:classifier}). The algorithm selected 100 SNPs, both from all regions mentioned above and new ones (S2C Fig). However, it was unclear why those SNPs were selected, as emphasized by the high P\=/value of some of them, which further complicates the biological interpretation. Moreover, and in opposition to what would be expected under the omnigenic model, the genes to which these SNPs map (Section~\ref{methods:snp2gene}) were not interconnected in the protein-protein interaction network (PPIN, Section~\ref{methods:networks}). Moreover, the classification performance of the model was low (sensitivity = 55\%, specificity = 55\%, Section~\ref{methods:benchmark}). Together, these issues motivated exploring network methods, which consider not only statistical association but also the location of each gene in a PPIN to find susceptibility genes.

@@

Topologically, the consensus consisted of a connected component composed of 49 genes and multiple smaller subnetworks (Fig~\ref{fig:consensus}B and \ref{fig:consensus}C). Among the latter, 19 genes were in subnetworks containing a single gene or two connected nodes. This implied that they did not have a consistently altered neighborhood but were strongly associated themselves and hence picked by at least two methods. The large connected component contained genes that are highly central in the PPIN. This property weakly anticorrelated with the P\=/value of association to the disease (Pearson correlation coefficient~=~-0.26, \nameref{sfig:consensus_stats}). This anticorrelation suggested that these genes were selected because they were on the shortest path between two high scoring genes. Because of this, we hypothesize that highly central genes might contribute to the heritability through alterations of their neighborhood, consistent with the omnigenic model of disease \cite{boyle_expanded_2017}. For instance, the most central node in the consensus solution was \emph{COPS5}, a component of the COP9 signalosome that regulates multiple signaling pathways. \emph{COPS5} is related to multiple hallmarks of cancer and is overexpressed in multiple tumors, including breast and ovarian cancer \cite{liu_jab1_cops5_2018}. Despite its lack of association in GENESIS or in studies conducted by the Breast Cancer Association Consortium (BCAC) \cite{Michailidou2017} (VEGAS2 P\=/value of 0.22 and 0.14, respectively), its neighbors in the consensus solution had consistently low P\=/values (median VEGAS2 P\=/value~=~0.006).

@@

We compared the results obtained with different network methods to the European sample of BCAC, the largest GWAS to date on breast cancer (Section~\ref{methods:bcac}). Although BCAC case-control studies do not necessarily target cases with a familial history of breast cancer like GENESIS does, this comparison is pertinent since we expect a shared genetic architecture at the gene level, at which most network methods operate. Together with BCAC's scale (90 times more samples than GENESIS), this shared genetic architecture provided a reasonable counterfactual of what we would expect if GENESIS had a larger sample size. We computed a gene association score on BCAC (Section~\ref{methods:bcac}). The solutions provided by the different  network methods overlapped significantly with BCAC hits (Fisher's exact test P\=/value \textless~0.019). The gene-based methods achieved comparable precision (2\%-25\%) and recall (1.3-12.1\%) at recovering BCAC-significant genes (S4A Fig). Interestingly, while SConES GI achieved a similar recall at the SNP-level (8.6\%), it showed a much higher precision (47.3\%).

@@

We compared the six network methods in a 5-fold subsampling setting (Section~\ref{methods:benchmark}). In this comparison we measured properties (Figs~\ref{fig:benchmark} and S4): the size of the solution; the sensitivity and the specificity of an L1-penalized logistic regression classifier on the selected SNPs; the stability of the methods; and their computational runtime. The solution size varies greatly between the different methods (Fig~\ref{fig:benchmark}A). Heinz produced the smallest solutions, with an average of 182 selected SNPs (Section~\ref{methods:snp2gene}) while the largest ones came from SConES GI (6\,256.6 SNPs) and dmGWAS (4\,255.0 SNPs). LEAN did not produce any solution in any of the subsamples.

To determine whether the selected SNPs could predict cancer susceptibility, we computed the classifiers' performances on \emph{test sets} (S4B Fig). The different classifiers displayed similarly low sensitivities and specificities, all in the 0.52 -- 0.56 range. Interestingly, the classifier trained on all the SNPs had a similar performance, despite being the only method aiming to minimize prediction error. Of course, although these performances were low, we did not expect to separate cases from controls well using exclusively genetic data \cite{deloscamposComplexTraitPredictionEra2018}.

@@

In the GENESIS dataset, heinz's solution was very conservative, providing a small solution with the lowest median P\=/value (Table~\ref{tab:gene_solutions}). By repeatedly selecting this compact solution, heinz was the most stable method (Fig~\ref{fig:benchmark}B). Its conservativeness stemmed from its preprocessing step, which modeled the gene P\=/values as a mixture model of a beta distribution and a uniform distribution, controlled by an FDR parameter. Due to the limited signal at the gene level in this dataset (S2B Fig), only 36 genes retained a positive score after that transformation. However, this small solution did not provide much insight into the susceptibility mechanisms to cancer. Importantly, it ignored genes that were associated with cancer in this dataset, like \emph{FGFR2}.

@@

We explored methods' parameter space by running them under different combinations of parameters (Section~\ref{methods:parameters}). In agreement with their formulations (Section~\ref{methods:methods}), larger values of specific parameters produced less stringent solutions (S7A Fig): for HotNet2 and heinz, this is the threshold above which genes receive a positive score; for dmGWAS, it is the \texttt{d} parameter, which controls how far neighbors could be added; for SigMod, it is \texttt{nmax}, which specifies the maximum size of the solution; and for LEAN, it is the P-value threshold to consider a solution significant. Two parameters had the opposite effect (the larger, the more stringent): SigMod's \texttt{maxjump}, which sets the threshold to consider an increment in $\lambda$ ``large enough''; and SConES' $\eta$, where higher values produce smaller solutions. However, two of the parameters did not have the expected effect: dmGWAS' \texttt{r}, which controls the minimum increment in the score required to add a gene; and SigMod's \texttt{maxjump}, which sets the threshold to consider an increment in $\lambda$ ``large enough''. In both cases, the size of the solution was very similar across the different values. Despite the differences in size, the solutions' size was relatively robust to the choice of parameters (Fig S7B).

@@

Most network methods, including the consensus, were highly unstable (Fig~\ref{fig:benchmark}B), raising questions about the results' reliability. We built a new, \emph{stable consensus} solution using the genes selected most often across the 30 solutions obtained by running the 6 methods on 5 different splits of the data (Section~\ref{methods:benchmark}). Such a network should capture the subnetworks more often found altered, and hence should be more resistant to noise. We used only genes selected in at least 7 solutions, which corresponded to 1\% of all genes selected at least once. The resulting stability-based consensus was composed of 68 genes (Fig~\ref{fig:stable-consensus}A). This network shared most of the properties of the consensus: breast cancer susceptibility genes were overrepresented (P\=/value = 3 \texttimes{} 10\textsuperscript{-4}), as well as genes involved in mitochondrial translation and the attenuation phase (adjusted P\=/values 0.001 and 3 \texttimes{} 10\textsuperscript{-5} respectively); the selected genes were more central than average (P\=/value = 1.1 \texttimes{} 10\textsuperscript{-14}); and a considerable number of nodes (19) were isolated (Fig~\ref{fig:stable-consensus}B and \ref{fig:stable-consensus}C).

@@
